---
title: 'Data Munging: Low-Income Energy Affordability Data'
author: "Eric Scheier"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  pdf_document: default
  md_document:
    variant: markdown_github
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(scales)
library(tidyverse)
library(readxl)
```
# Introduction

```{r params}
energy_poverty_line <- 0.10
acs_date <- 2016
youngest_building <- acs_date - 1
oldest_building <- 1900 # this assumption is not used in the ultimate analysis
max_units <- 500 # 1 NYC block is 5 acres, assume 100 units/acre # this assumption is not used in the ultimate analysis
```


This is a data munging report for the Low Income Energy Affordability Data [LEAD](https://catalog.data.gov/dataset/low-income-energy-affordability-data-lead-tool) from the Department of Energy (DOE). This dataset represents an estimate of the monthly energy bills and average incomes of households in the united states, segmented by the following characteristics:

+ Year of building first construction
+ Number of units in the building
+ Primary heating fuel type
+ Whether the residents own or rent the unit
+ Household income relative to Area Median Income or Federal Poverty Level
+ State
+ County, City, or Census Tract

For this project, I will use the data which displays the results by census tract in terms of Area Median Income for the state of North Carolina.

# Acquiring Data

```{r cache=TRUE}
base_url <- "https://openei.org/doe-opendata/dataset/9dcd443b-c0e5-4d4a-b764-5a8ff0c2c92b/resource/831ac481-1a60-4f87-a887-15b310ffff53/download/"


file_name <- "ami68tractshnc"
data_url <- paste0(base_url,file_name,".csv")
desired_name <- paste0(file_name,".csv")
desired_path <- file.path(desired_name)

if (!file.exists(desired_path)){
  download.file(data_url,desired_path)
}
```

A full data dictionary is available. It is downloaded and loaded into constituent dataframes below.

```{r}
# https://readxl.tidyverse.org/articles/articles/readxl-workflows.html

read_then_csv <- function(sheet, path) {
  pathbase <- path %>%
    basename() %>%
    tools::file_path_sans_ext()
  path %>%
    read_excel(sheet = sheet, col_names = FALSE, col_types = "text") %>% 
    write_csv(paste0(sheet, ".csv"))
}

url <- "https://openei.org/doe-opendata/dataset/9dcd443b-c0e5-4d4a-b764-5a8ff0c2c92b/resource/51a2cd49-fd61-4842-82e2-2f90ffec7e42/download/datadictionary.xlsx"

if(!file.exists("Data Dictionary.csv")){
  temp <- tempfile()
  download.file(url,temp,mode="wb")
  path <- temp
  path %>%
    excel_sheets() %>%
    set_names() %>% 
    map(read_then_csv, path = path)
}

data_dict <- read_csv("Data Dictionary.csv")

ybl_dict <- data_dict[1:7,1:2]
ybl_dict[1,2] <- data_dict[1,3]
names(ybl_dict) <- as.matrix(ybl_dict[1, ])
ybl_dict <- ybl_dict[-1, ]

bld_dict <- data_dict[9:18,1:2]
bld_dict[1,2] <- data_dict[9,3]
names(bld_dict) <- as.matrix(bld_dict[1, ])
bld_dict <- bld_dict[-1, ]

hfl_dict <- data_dict[20:29,1:2]
hfl_dict[1,2] <- data_dict[20,3]
names(hfl_dict) <- as.matrix(hfl_dict[1, ])
hfl_dict <- hfl_dict[-1, ]

burden_dict <- data_dict[1:9,7:8]
names(burden_dict) <- c("Variable", "Description")
burden_dict <- rbind(burden_dict, 
                     names(hfl_dict), 
                     names(bld_dict), 
                     names(ybl_dict))
burden_dict
```


```{r}
data <- read_csv(desired_path)

summary(data)
```

# Data Munging

Munging Steps:

+ Seperate the `AMI68` column into `occupancy_type` and `income_bracket` columns.
+ Create `min_age` from `YBL INDEX`.
+ Create `min_units` from `BLD INDEX`.
+ Create `detached` from `BLD INDEX`.
+ Create the Energy Expenditures Indicator `mean_energy_cost`.
+ Create the Energy Burden Indicator `mean_energy_burden`.
+ Remove rows where `UNITS<1`, `mean_energy_cost<0`, or `HINCP<0`.

## Break Out Ownership & Income Bracket

The data combines the unit's ownership status (`OWNER` vs. `RENTER`) and income bracket as a fraction of Area Median Income (`0-30%`, `30-60%`, `60-80%`, `80-100%`, or `100%+`) into the same column (e.g. `r data$AMI68[1]`). This `AMI68` column needs to be seperated for meaningful analysis. These categorical variables are saved as factors.

```{r}
#Seperate the `AMI68` column into `occupancy_type` and `income_bracket` columns

data <- data %>% 
  separate(col="AMI68",
           into=c("occupancy_type", "income_bracket"), 
           sep = " ", 
           remove = FALSE, 
           convert = FALSE,
           extra = "warn", 
           fill = "warn")

data$occupancy_type <- as.factor(data$occupancy_type)
data$income_bracket <- as.factor(data$income_bracket)
```

## Create Meaningful Indicators from Indices

The `YBL INDEX` and `BLD INDEX` columns need some further treatment. `YBL INDEX` represents the year the building was first constructed, but is an index seperated into 10-20 year increments dating back to 1940. 

```{r}
ybl_dict
```


Since these indeces represent different possible ages and the increments are not uniform, I will add a numerical indicator (`min_age`) to represent the youngest age the building could possibly be based on its category. Thus, a building in the `2010+` range must be at least 0 years old (this data was collected in `r acs_date`), in the `1980-99` range at least `r acs_date-1980` years old, in the `BEFORE 1940` range at least `r acs_date-1940`, etc. I choose youngest rather than oldest age because it avoids the requirement to make an assumption about the oldest age of buildings built before 1940.

For simplicity, I assume that the youngest building is 1 year old.

```{r}
#Create `min_age` from `YBL INDEX`.

ybl_ranges <- str_extract_all(ybl_dict$`Year of building first construction`, "[0-9]+", simplify=TRUE)
ybl_ranges <- apply(ybl_ranges, c(1,2), as.numeric)
ybl_ranges[1,2] <- youngest_building
ybl_ranges[6,2] <- ybl_ranges[6,1]
ybl_ranges[6,1] <- oldest_building
ybl_ranges[2:5,2] <- floor(ybl_ranges[2:5,1]/100)*100 + ybl_ranges[2:5,2]

ybl_ranges <- data.frame(ybl_ranges)
names(ybl_ranges) <- c("min_year", "max_year")
ybl_ranges$min_age <- acs_date - ybl_ranges$max_year
ybl_ranges$max_age <- acs_date - ybl_ranges$min_year
ybl_dict <- cbind(ybl_dict, ybl_ranges)
ybl_dict$`YBL INDEX` <- as.factor(ybl_dict$`YBL INDEX`)


ybl_dict <- rename(ybl_dict,
                   year_constructed=`Year of building first construction`)

ybl_dict$year_constructed <- as.factor(ybl_dict$year_constructed)

data$`YBL INDEX` <- as.factor(data$`YBL INDEX`)
data <- merge(data, ybl_dict[c("YBL INDEX",
                               "min_age",
                               "year_constructed")], 
              by = "YBL INDEX", 
              all.x = TRUE)
```

Similar to building age, the variable `BLD INDEX` represents a non-uniformly distributed set of buckets for the range of `number of units in the building`, as well as whether single unit households are attached or detached from neighboring households. I will extract the minimum number of units from the range, and whether the building is detached.

Those households labeled `OTHER UNIT` will be given values of `NA` for this characteristic.

```{r}
#Create `min_units` from `BLD INDEX` (and `max_units`).
#Create `detached` from `BLD INDEX`.

bld_ranges <- str_extract_all(bld_dict$`Number of units in the building`, "[0-9]+", simplify=TRUE)

bld_ranges <- apply(bld_ranges, c(1,2), as.numeric)
bld_ranges[1:3,2] <- bld_ranges[1:3,1]
bld_ranges[8,2] <- max_units

bld_ranges <- data.frame(bld_ranges)
names(bld_ranges) <- c("min_units", "max_units")

bld_ranges$detached <- as.factor(c(1,rep(0,nrow(bld_ranges)-2),NA))

bld_dict <- cbind(bld_dict, bld_ranges)
bld_dict$`BLD INDEX` <- as.factor(bld_dict$`BLD INDEX`)

bld_dict <- rename(bld_dict,
                   number_of_units=`Number of units in the building`)

bld_dict$number_of_units <- as.factor(bld_dict$number_of_units)

data$`BLD INDEX` <- as.factor(data$`BLD INDEX`)
data <- merge(data, bld_dict[c("BLD INDEX",
                               "min_units",
                               "detached",
                               "number_of_units")], 
              by = "BLD INDEX", 
              all.x = TRUE)
```

## Assign Primary Heating Fuel Type

```{r}
hfl_dict <- rename(hfl_dict,
               primary_heating_fuel=`Primary heating fuel type`)

hfl_dict$primary_heating_fuel <- as.factor(hfl_dict$primary_heating_fuel)

data <- merge(data, hfl_dict[c("HFL INDEX","primary_heating_fuel")], 
              by = "HFL INDEX", 
              all.x = TRUE)
```



## Create the Energy Burden Indicators

Creating the final metric of average energy burden for each cohort is as simple as adding up the montly expenditures on electricity (`ELEP CAL`), natural gas (`GASP CAL`), and other fuels (`FULP`), and dividing this by the cohort's average monthly income (`HINCP / 12.0`).

```{r}
# Create the Energy Expenditures Indicator `mean_energy_cost`.

data$mean_energy_cost <- data$`ELEP CAL` + data$`GASP CAL` + data$FULP

# Create the Energy Burden Indicator `mean_energy_burden`.

data$mean_energy_burden <- data$mean_energy_cost / (data$HINCP/12.0)

```

## Clean-up

+ rename variables for convenience
+ keep only what we need

Select the columns I want:

```{r}
data <- select(data,
               `GEO ID`,
               `PUMA10`,
               `FMR`,
               `occupancy_type`,
               `income_bracket`,
               `primary_heating_fuel`,
               `number_of_units`,
               `year_constructed`,
               `UNITS`, 
               `HINCP`, 
               `ELEP CAL`, 
               `GASP CAL`, 
               `FULP`, 
               `COUNT`, 
               `min_age`, 
               `min_units`, 
               `detached`,
               `mean_energy_cost`,
               `mean_energy_burden`)
```

Rename columns to remove spaces:

```{r}
data <- rename(data,
               geo_id=`GEO ID`,
               puma10_code=`PUMA10`,
               fmr_code=`FMR`,
               households=`UNITS`, 
               acs_responses=`COUNT`,
               annual_income=`HINCP`,
               electricity_spend=`ELEP CAL`, 
               gas_spend=`GASP CAL`, 
               other_spend=`FULP`)
```

Decompose the geo_id into state, county, and tract codes for interpretibility.

```{r}
data <- separate(data,
                 col=geo_id,
                 into=c("state_id","county_id","tract_id"),
                 sep=c(2,5), 
                 remove=FALSE, 
                 convert=TRUE)
```

## Adding an Energy Poverty Response Variable

For further analysis, I will add a designation of whether a cohort is, on average, in energy poverty depending on whether the mean energy burden is above `r label_percent()(energy_poverty_line)`.

```{r}
data$energy_poverty <- as.logical(data$mean_energy_burden > energy_poverty_line)
```

## Remove Fractional Units

The estimation procedure used by the DOE results in an estimated number of occupied housing units meeting the subset characteristics (`UNITS`, renamed as `households`) and displays the number of American Community Survey responses that contribute to the estimate of energy expenditures (`COUNT`, renamed as `acs_responses`). We first remove any categories that have fewer than 1 unit represented, since this is not physically possible.

This results in removing `r 100*sum(data$households<1)/length(data$households)`% of the available rows (gross: `r sum(data$households<1)` rows of `r length(data$households)`). This is a total of `r sum(data$households[data$households<1])` housing units, or `r 100*sum(data$households[data$households<1])/sum(data$households)`% of the estimated total `r sum(data$households)` units in the state.

```{r}
weird_rows <- data$annual_income<=0 | data$mean_energy_cost<=0
# | data$households<1 "or those which represent les than a single housing unit "
# | data$mean_energy_burden>1 "or energy costs greater than income"
weird_data <- data[weird_rows,]
data <- data[!weird_rows,]

weird_rate <- nrow(weird_data)/(nrow(data)+nrow(weird_data))
weird_unit_rate <- sum(weird_data$households)/(sum(data$households)+sum(weird_data$households))
```

I have seperated any cohorts with incomes or energy costs less than or equal to $0 into another dataset for analysis. This represents approximately `r round(100*weird_unit_rate,0)`% of the housing units and `r round(100*weird_rate,0)`% of the examined cohorts, so it will be important to make sure that this subset does not contain systematic bias.

```{r}
write_csv(weird_data, "weird_lead.csv")
write_csv(data,"clean_lead.csv")
```

```{r}
data_reloaded <- read_csv(file.path(getwd(),"clean_lead.csv"),
                          col_types=cols(
                            geo_id = readr::col_factor(),
                            state_id = readr::col_factor(),
                            county_id = readr::col_factor(),
                            tract_id = readr::col_factor(),
                            puma10_code = readr::col_factor(),
                            fmr_code = readr::col_factor(),
                            occupancy_type = readr::col_factor(),
                            income_bracket = readr::col_factor(),
                            primary_heating_fuel = readr::col_factor(),
                            number_of_units = readr::col_factor(),
                            year_constructed = readr::col_factor(),
                            households = col_double(),
                            annual_income = col_double(),
                            electricity_spend = col_double(),
                            gas_spend = col_double(),
                            other_spend = col_double(),
                            acs_responses = col_double(),
                            min_age = col_double(),
                            min_units = col_double(),
                            detached = readr::col_factor(),
                            mean_energy_cost = col_double(),
                            mean_energy_burden = col_double(),
                            energy_poverty = col_logical()
                            ))

all(data_reloaded == data, na.rm=TRUE)
summary(data_reloaded)
```

